Assessment of the ability to track changes for measurements, nowcasts, and forecasts

### 2024-04-11 paper.tex
Update on Overleaf.
Shorten Paper

@@

% \subsection{Computing changes and notation}\label{subsec:notation}

As noted above, the assessment of \ac{atc} is based on the observed and predicted changes over a time horizon $l$.
The \textit{observed} change is straightforward to compute for all types of \ac{mnf}.
Let $\mathbf{y} = (y_t)_{t=0}^T$ denote the true values for nowcasting or forecasting, or gold standard measurements up to time $T$.
The sequence of observed changes is then given by the differences of values in $\mathbf{y}$ with horizon $l$, that is,
\begin{equation}\label{eq:diffy}
    \diffylt = (y_{t} - y_{t-l}) \quad \text{for}\ t = l, \dots, T.
\end{equation}

@@

The notation is similar for forecasting: Let $x_{t | \tau}$ denote the forecast for target time $t$ and issue time $\tau$.
The predicted change is computed by
\begin{equation} \diffxl = (x_{t|t-l} - y_{t-l})^T_{t=l}\label{eq:diffxl_forecasting}
\end{equation}
with the same structure as in the nowcasting case.
The indices are chosen to be consistent with $\diffyl$.
If the true value $y_{t-l}$ is not known at time $t-l$, a similar modification as in Equation~\eqref{eq:diffxl_nowcasting} can be made.

@@

In applications, data are often not available for all time steps, for example, due to technical problems or delays in data transfer (see the examples in Sections~\ref{sec:application-covid} and~\ref{sec:application-eda}).
We refer to time steps for which either \ac{mnf} or true values, or both, are unavailable as missing values.
Systematical missing values could lead to a biased assessments, and missing data should be inspected for any underlying patterns.
If the missing values are not systematic, random, and occur scarcely, data pairs with missing values can be excluded from the data to calculate the measures~\parencite[see][Section 1.3]{VanBuuren2018}.
Note that in the case of measurement data, one missing value in the time series leads to two undefined differences in the change series; that is if $x_t$ is missing, $\diffxt$ and $\diffxt[t+l]$ are undefined; if an observation $y_t$ is missing, $\diffyt$ and $\diffyt[t+l]$ are undefined.
The data pair is excluded even if the corresponding nowcast or forecast is available.

@@

Formally, the assessment of \ac{atc} is the same for \ac{mnf}, given the notation for the respective application of Section~\ref{subsec:notation}.
In the following, we omit the horizon $l$ for ease of notation; $\diffx$ and $\diffy$ refer to $\diffxl$ and $\diffyl$ for a common horizon $l$.
The \ac{atc} is maximal if all the change directions are predicted correctly; that is, the sign of all elements of $\diffx$ and $\diffy$ coincide.
Consequently, when assessing the \ac{atc}, we examine the statistical consistency of $\sign(\diffx)$ and $\sign(\diffy)$.
A simple yet insightful method is the four-quadrant plot~\citep[see, e.g., ][]{Perrino1998,Saugel2015}.
In a four-quadrant plot, the occured changes are plotted with the predicted changes, that is, $(\diffyt, \diffxt)$ for $t = l, \dots, T$.
Thus, the x-axis of a four-quadrant plot shows the true value differences, whereas the y-axis displays the prediction data differences.
Points in the green upper right and lower left quadrants reflect a correct change direction for the respective time step, whereas points in the remaining red quadrants show incorrect predicted changes.
Figure~\ref{fig:aatc_basic_4q} displays a basic four-quadrant plot and
Figure~\ref{fig:aatc_basic_4q_sample} shows a four-quadrant graph for simulated data with $T=1461$, for example, four years of daily data.

It can be extended by including information on the time index in the point color to reveal effects over time.
In Figure~\ref{fig:aatc_basic_4q_sample_color}, the point colors turn from blue to green for higher time indices $t$, that is, more recent values; $(\diffyt[l], \diffxt[l])$ is blue and turns green until $(\diffyt[T], \diffxt[T])$.
However, four-quadrant plots become crowded for larger datasets, and sequential information on the differences is complex to assess thoroughly.

@@

\begin{figure}
\centering
\begin{subfigure}[t]{.24\textwidth}
\includegraphics{plots/illustrative_examples/4Q_without_excl}
\caption{Basic four-quadrant plot.} \label{fig:aatc_basic_4q}
\end{subfigure}\hspace{0.01\textwidth}%
\begin{subfigure}[t]{.24\textwidth}
\includegraphics{plots/illustrative_examples/4q_excl_box}
\caption{Four-quadrant plot with rectangular exclusion area.}\label{fig:aatc_basic_4q_excl_box}
\end{subfigure}\hspace{0.01\textwidth}%
\begin{subfigure}[t]{.24\textwidth}
\includegraphics{plots/illustrative_examples/4q_excl_axis}
\caption{Four-quadrant plot with horizontal exclusion area.} \label{fig:aatc_basic_4q_excl_axis}
\end{subfigure}\hspace{0.01\textwidth}%
\begin{subfigure}[t]{.24\textwidth}
\includegraphics{plots/illustrative_examples/4q_excl_cross}
\caption{Four-quadrant plot with cross-shaped exclusion area.}\label{fig:aatc_basic_4q_excl_cross}
\end{subfigure}%
\caption{Illustrations of the four-quadrant plot with sample points and with and without exclusion areas. In Figure~\ref{fig:aatc_basic_4q}, points 2, 3, 5, and 6 correctly track change, whereas points 1, 4, and 7 do not.}
\label{fig:aatc_4q}
\end{figure}

@@

Analyzing the number of points in the green versus red quadrants is a standard approach in the \ac{atc} assessment of measurement data~\citep{Critchley2010,Saugel2015}.
With that, we mainly estimate the probability of a correctly predicted change direction, $P(\diffxrv \diffyrv > 0)$, where $\diffyrv$ and $\diffxrv$ denote random variables for future incremental changes.
Since $z_1 z_2 > 0$ imposes the same condition as $\sign(z_1) = \sign(z_2)$ ($z_1, z_2 \in \R \setminus \{ 0 \}$), the standard estimator for $P(\diffxrv \diffyrv > 0)$ is
\begin{equation}
    \acc (\diffx, \diffy) \coloneqq \frac{\sum_{t \in \mathcal{T}} \ind{\diffxt \diffyt > 0}}{\card{T}}.\label{eq:acc}
\end{equation}
Here, the numerator counts the number of same-sign-changes, while the denominator is the number of considered pairs $(\diffyt, \diffxt)$.
Thus, $\acc$ is the proportion of concordant changes on all changes.
We refer to this estimator as the \textit{ATC ratio} of the prediction and set $\mathcal{T} = \{l, \dots, T\}$.
Visually, the measure computes the fraction of points in the upper right or lower left quadrant.
Similar evaluations are used in other scientific areas, for example, with contingency tables in dichotomous forecasting or with confusion matrices in classification analysis~\citetext{see, e.g., the introductions in \citealt{James2021}, Ch. 4, and \citealp{Jolliffe2012}, Ch. 3}.
Many other measures can be adapted from those fields to deepen the analysis.
Two simple measures that focus on either a positive or negative predicted change are the positive and negative \ac{atc} ratios $\accp$ and $\accm$, respectively.
They are defined as
\begin{align}
    \accp (\diffx, \diffy) &\coloneqq \frac{\sum_{t \in \mathcal{T}} \ind{\diffxt \diffyt > 0} \ind{\diffxt > 0}}{\sum_{t \in \mathcal{T}} \ind{\diffxt > 0}},\ \text{and} \label{eq:accp}\\
    \accm (\diffx, \diffy) &\coloneqq \frac{\sum_{t \in \mathcal{T}} \ind{\diffxt \diffyt > 0} \ind{\diffxt < 0}}{\sum_{t \in \mathcal{T}} \ind{\diffxt < 0}}.\label{eq:accm}
\end{align}
They estimate the probability of a correct prediction of the direction of change, given that the predicted direction is positive or negative, that is, $P(\diffxrv \diffyrv > 0 | \diffxrv > 0)$ and $P(\diffxrv \diffyrv > 0 | \diffxrv < 0)$.

@@

\begin{figure}
    \centering
    \begin{subfigure}[t]{.24\textwidth}
\includegraphics{plots/illustrative_examples/4Q_sample_without_time}
\caption{Four-quadrant plot with simulated data.}\label{fig:aatc_basic_4q_sample}
\end{subfigure}\hspace{0.01\textwidth}
\begin{subfigure}[t]{.24\textwidth}
\includegraphics{plots/illustrative_examples/4Q_sample_with_time}
\caption{The same data is colored according to the time index $t$, the greener, the later.}\label{fig:aatc_basic_4q_sample_color}
\end{subfigure}\hspace{0.01\textwidth}
\begin{subfigure}[t]{.48\textwidth}
    \includegraphics{plots/illustrative_examples/trending_ratio_time_series}
    \caption{Rolling estimate of the \ac{atc} ratio over time with window length 91. }\label{fig:atc_ratio_time_series}
    \end{subfigure}%
    \caption[Visualizations of the four-quadrant plot for data with a time-varying ATC ratio.]{Visualizations for data with a time-varying ATC ratio. We defer information on the data generation process to the appendix (see \ref{subsec:app-ATC-data-generation}). The ATC ratio for the entire data set is $\mu = 0.7577$. The strong seasonality of the ATC ratio becomes visible in Figure~\ref{fig:atc_ratio_time_series}. The green curve $k_t$ shows the theoretical probability that $\diffxt$ has the same sign as $\diffyt$ for each time step. The rolling estimates are shifted to the right compared to $k_t$ as the windows look backward. Thus, the yearly course of the ATC ratio can be detected.
The ATC ratio has a strong sinus-shaped seasonality with a peak after a quarter of a year and a low point after three quarters.
}
\end{figure}

@@

The above measures can be extended to account for information on the point's location within the quadrant.
For example, points close to the zero point may have less explanatory power or may be less reliable than points far away from zero on one of the diagonals.
Suppose noise or non-systematic effects are present in the true values or predictions.
In that case, a point's assignment to a quadrant can be driven by noise instead of a systematic \ac{atc}.
This is more likely for points with at least one small coordinate.

Using an exclusion area around the zero point, as further defined below, is a straightforward and highly interpretable extension of the measures of Section~\ref{subsec:atc-measures} accounting for such effects~\citep[see, e.g.,][]{Saugel2015,Critchley2010}.
Points within that area are omitted in the four-quadrant plot and the calculation of the measures.
In particular, the prediction models are likely to have a noise component; thus, $\diffx$ should be subject to an exclusion area.
We denote the measures of Equations~\eqref{eq:acc},~\eqref{eq:accp} and~\eqref{eq:accm} accounting for an exclusion area $E$ by
\begin{align}
    \acceps (\diffx, \diffy, E) &\coloneqq \frac{\sum_{t \in \mathcal{T}} \ind{\diffx \diffy > 0} \ind{(\diffyt, \diffxt) \notin E}}{\sum_{t \in \mathcal{T}} \ind{(\diffyt, \diffxt) \notin E}},\label{eq:acceps}\\
    \accpeps (\diffx, \diffy, E) &\coloneqq \frac{\sum_{t \in \mathcal{T}} \ind{\diffxt \diffyt > 0} \ind{\diffxt > 0, (\diffyt, \diffxt) \notin E}}{\sum_{t \in \mathcal{T}} \ind{\diffxt > 0, (\diffyt, \diffxt) \notin E}}, \ \text{and} \label{eq:accpeps}\\
    \accmeps (\diffx, \diffy, E) &\coloneqq \frac{\sum_{t \in \mathcal{T}} \ind{\diffxt \diffyt > 0} \ind{\diffxt < 0, (\diffyt, \diffxt) \notin E}}{\sum_{t \in \mathcal{T}} \ind{\diffxt < 0, (\diffyt, \diffxt) \notin E}}. \label{eq:accmeps}
\end{align}
The additional argument $(\diffyt, \diffxt) \notin E$ in the indicator ensures that only points not lying in the exclusion are evaluated when computing the measures.
The measures are then estimators for the probability of predicting the correct direction, given that the point's location is not driven by noise or non-informative changes.

The estimators accept various shapes of the exclusion area.
Figure~\ref{fig:aatc_4q} visualizes examples of exclusion area shapes.
A rectangular exclusion area, $E = \{(x, y) \in \R^2: (-\varepsilon_x \leq x \leq \varepsilon_x) \land (-\varepsilon_y \leq y \leq \varepsilon_y) \}$ for $\varepsilon_x, \varepsilon_y > 0$, leaves out points that are small in both components and, therefore, likely to be driven by noise.
Points are included if at least one of true value or prediction is unlikely to be noise-driven.
In the example graph in Figure~\ref{fig:aatc_basic_4q_excl_box}, only point 1 is excluded.
An exclusion area along one axis, for example, $E = \{(x, y) \in \R^2: (-\varepsilon_x \leq x \leq \varepsilon_x)\}$ for $\varepsilon_x > 0$, removes points in which one of the components could change sign by a small amount of noise.
This particularly suits \ac{mnf} where small amounts of noise are inevitable.
A cross-shaped exclusion area, $E = \{(x, y) \in \R^2: (-\varepsilon_x \leq x \leq \varepsilon_x) \lor (-\varepsilon_y \leq y \leq \varepsilon_y) \}$ for $\varepsilon_x, \varepsilon_y > 0$, along both axes accounts for the sign reversal in both components.
A point is only included if no component is likely to be driven by noise.
For the latter two methods, points 1 and 7 in Figure~\ref{fig:aatc_basic_4q_excl_axis} or 1, 6, and 7 are excluded in Figure~\ref{fig:aatc_basic_4q_excl_cross}, respectively.

@@

Confidence intervals can account for the estimation uncertainty of the measures above.
Bootstrap confidence intervals are based on resampling and not on parametric assumptions as classical confidence intervals are~\parencite[for introductions see][]{Hesterberg2011,Bittmann2021}.
Many new samples are drawn with replacement from the dataset, and the statistic of interest is computed for each sample, yielding an estimate for the distribution of the statistic of interest.
Based on the derived \enquote{new} samples of the statistic, the confidence intervals can be derived through differently sophisticated bootstrapping methods.
We use the \ac{bca} approach for bootstrapping in the following, as it holds the confidence level for small and large samples and has a moderate computation time (see the simulation study in Appendix~\ref{subsec:appendix-aatc-bootstrap}).

@@

The estimators described above give information on the probabilities $P(\diffxrv \diffyrv > 0 | \diffxrv \diffyrv \notin E)$, $P(\diffxrv \diffyrv > 0 | \diffxrv > 0, \diffxrv \diffyrv \notin E)$ and $P(\diffxrv \diffyrv > 0 | \diffxrv < 0, \diffxrv \diffyrv \notin E)$.
A still finer analysis might be gained by considering the conditional distribution $P(\diffxrv \diffyrv > 0 | \diffxrv = \xcond)$ to assess the \ac{atc} ability of a prediction for a certain change $\diffxrv = \xcond$ of the \ac{mnf}.
Thereby, $P(\diffxrv \diffyrv > 0 | \diffxrv = \xcond)$ denotes the probability of a correct direction given a predicted change of $\xcond$.
Thus, if a change of $\xcond$ is observed in practice, one can directly assess its credibility.
A multivariate \acf{kde} facilitates a continuous estimation of $P(\diffxrv \diffyrv > 0 | \diffxrv = \xcond)$ by estimating the components $f_{\diffxrv, \diffyrv}$ and $f_{\diffxrv}$ of
\begin{align*}
P(\diffxrv \diffyrv > 0 | \diffxrv = \xcond) = \begin{cases}
                                              \int_{-\infty}^0 \frac{f_{\diffxrv, \diffyrv}(\xcond, y)}{f_{\diffxrv}(\xcond)} \ \textrm{d} \: y & \text{if } \xcond < 0, \\
                                              \int_{0}^{\infty} \frac{f_{\diffxrv, \diffyrv}(\xcond, y)}{f_{\diffxrv}(\xcond)} \ \textrm{d} \: y & \text{if } \xcond > 0, \\
\end{cases}
\end{align*}
for $\xcond \neq 0$ through a \ac{kde}.
A comprehensive introduction to multivariate \ac{kde} can be found in \textcite{Gramacki2018}, and implementations are available in many programming languages, for example, in the \verb|statsmodels| in Python~\parencite{Seabold2010}.
The \ac{kde} yields estimates for $P(\diffxrv \diffyrv > 0 | \diffxrv = \xcond)$ for all values of $\xcond \in \R$.
Multivariate \ac{kde} takes a kernel and bandwidth selector as modeling parameters.
We advise using a Gaussian kernel and the cross-validation maximum likelihood as bandwidth selector (see Appendix~\ref{subsec:appendix-kde}).

Assessing $P(\diffxrv \diffyrv > 0 | \diffxrv = \xcond)$ graphically by drawing $P(\diffxrv \diffyrv > 0 | \diffxrv = \xcond)$ against $\xcond$ eases the simultaneous evaluation of various $\xcond$.
Furthermore, the graph facilitates the comparison of various methods in a single graph, and asymmetries of $P(\diffxrv \diffyrv > 0 | \diffx = \xcond)$ with respect to $x$ in the \ac{atc} can be detected.
We refer to the plot as a \textit{conditional ATC plot}.

@@

In nowcasting and forecasting, probabilistic predictions have become more prevalent in recent years (see, for example, Sections~\ref{sec:application-covid} and~\ref{sec:application-eda} or the review in~\cite{Gneiting2014}).
In this section, we develop \ac{atc} assessments for probabilistic measurements and nowcasts.
Probabilistic predictions issue a probability distribution for the quantity of interest based on their available information and, thus, include a point estimate and information on the prediction uncertainty and quantiles simultaneously.
For \ac{atc} analysis, probabilistic predictions thus also contain a probability of a positive or negative change.
In this section, we extend the \ac{atc} assessment to probabilistic predictions.

Probabilistic predictions can take the form of a \ac{cdf}, \ac{pdf}, or quantiles.
The \ac{cdf} is the most general and can be used to derive the others, given they exist.
Let us first assume that the prediction is a \ac{cdf}, and that $y_{t-l}$ is known at time $t$ (see Table~\ref{tab:notation}).
In Appendix~\ref{subsec:app-probabilistic-atc-evaluation}, we extend the analysis for predictions given as quantiles or unknown true values.

>>>>

% (Approximate) Word Count
% Abstract

@@

\subsection{Computing changes and notation}\label{subsec:notation}

As noted above, the assessment of \ac{atc} is based on the observed and predicted changes over a time horizon $l$.
The \textit{observed} change is straightforward to compute for all types of \ac{mnf}.
Let $\mathbf{y} = (y_t)_{t=0}^T$ denote the actual values for nowcasting or forecasting, or gold standard measurements up to time $T$.
The sequence of observed changes is then given by the differences of values in $\mathbf{y}$ with horizon $l$, that is,
\begin{equation}\label{eq:diffy}
    \diffylt = (y_{t} - y_{t-l}) \quad \text{for}\ t = l, \dots, T.
\end{equation}

@@

The notation is similar for forecasting: Let $x_{t | \tau}$ denote the forecast for target time $t$ and issue time $\tau$.
The predicted change is computed by
\begin{equation} \diffxl = (x_{t|t-l} - y_{t-l})^T_{t=l}\label{eq:diffxl_forecasting}
\end{equation}
with the same structure as in the nowcasting case and consistent indices with $\diffyl$.
If the true value $y_{t-l}$ is not known at time $t-l$, a similar modification can be made as in Equation~\eqref{eq:diffxl_nowcasting}.

@@

In applications, data are often not available for all time steps, for example, due to technical problems or delays in data transfer (see the examples in Sections~\ref{sec:application-covid} and~\ref{sec:application-eda}).
We refer to time steps for which either \ac{mnf} or true values are unavailable as missing values.
Systematical missing values could lead to a biased assessment, and missing data should be inspected for any underlying patterns.
If the missing values are not systematic, random, and occur scarcely, data pairs with missing values can be excluded from the data to calculate the measures~\parencite[see][Section 1.3]{VanBuuren2018}.
Note that in the case of measurement data, one missing value in the time series leads to two undefined differences in the change series; that is if $x_t$ is missing, $\diffxt$ and $\diffxt[t+l]$ are undefined; if an observation $y_t$ is missing, $\diffyt$ and $\diffyt[t+l]$ are undefined.
The data pair is excluded even if the corresponding nowcast or forecast is available.

@@

Formally, the assessment of \ac{atc} is the same for \ac{mnf}, given the notation for the respective application of Section~\ref{subsec:notation}.
In the following, we omit the horizon $l$ for ease of notation; $\diffx$ and $\diffy$ refer to $\diffxl$ and $\diffyl$ for a common horizon $l$.
The \ac{atc} is maximal if all predicted change directions are correct; that is, the sign of all elements of $\diffx$ and $\diffy$ coincide.
Consequently, when assessing the \ac{atc}, we examine the statistical consistency of $\sign(\diffx)$ and $\sign(\diffy)$.
A simple yet insightful method is the four-quadrant plot~\citep[see, e.g., ][]{Perrino1998, Saugel2015}.
In a four-quadrant plot, the occured changes and the predicted changes are plotted together, that is, $(\diffyt, \diffxt)$ for $t = l, \dots, T$.
Thus, the x-axis of a four-quadrant plot shows the true value differences, whereas the y-axis displays the prediction data differences.
Points in the green upper right and lower left quadrants reflect a correct change direction for the respective time step, whereas points in the remaining red quadrants show incorrectly predicted changes.
Figure~\ref{fig:aatc_basic_4q} displays a basic four-quadrant plot and Figure~\ref{fig:aatc_basic_4q_sample} shows a four-quadrant graph for simulated data with $T=1461$, for example, four years of daily data.

The four-quadrant plot can be extended by including information on the time index in the point color to reveal effects over time.
In Figure~\ref{fig:aatc_basic_4q_sample_color}, the point colors turn from blue to green for higher time indices $t$, that is, more recent values; $(\diffyt[l], \diffxt[l])$ is blue and turns green until $(\diffyt[T], \diffxt[T])$.
However, four-quadrant plots become crowded for larger datasets, and sequential information on the differences is complex to assess thoroughly.

@@

\begin{figure}
\centering
\begin{subfigure}[t]{.24\textwidth}
\includegraphics{plots/illustrative_examples/4Q_without_excl}
\caption{Basic four-quadrant plot.} \label{fig:aatc_basic_4q}
\end{subfigure}\hspace{0.01\textwidth}%
\begin{subfigure}[t]{.24\textwidth}
\includegraphics{plots/illustrative_examples/4q_excl_box}
\caption{Four-quadrant plot with rectangular exclusion area. Point 1 is excluded.}\label{fig:aatc_basic_4q_excl_box}
\end{subfigure}\hspace{0.01\textwidth}%
\begin{subfigure}[t]{.24\textwidth}
\includegraphics{plots/illustrative_examples/4q_excl_axis}
\caption{Four-quadrant plot with horizontal exclusion area. Points 1 and 7 are excluded.} \label{fig:aatc_basic_4q_excl_axis}
\end{subfigure}\hspace{0.01\textwidth}%
\begin{subfigure}[t]{.24\textwidth}
\includegraphics{plots/illustrative_examples/4q_excl_cross}
\caption{Four-quadrant plot with cross-shaped exclusion area. Points 1, 6 and 7 are excluded.}\label{fig:aatc_basic_4q_excl_cross}
\end{subfigure}%
\caption{Illustrations of the four-quadrant plot with sample points and with and without exclusion areas. The rectangular exclusion area in Figure~\ref{fig:aatc_basic_4q_excl_box} excludes only points where both components are likely to be noise-driven, while the exclusion areas in Figures~\ref{fig:aatc_basic_4q_excl_axis} and~\ref{fig:aatc_basic_4q_excl_cross} excludes points where at least one component is noise-driven. }
\label{fig:aatc_4q}
\end{figure}

@@

Analyzing the number of points in the green versus red quadrants is a standard approach in the \ac{atc} assessment of measurement data~\citep{Critchley2010, Saugel2015}.
With that, we estimate the probability of a correctly predicted change direction, $P(\diffxrv \diffyrv > 0)$, where $\diffyrv$ and $\diffxrv$ denote random variables for future incremental changes.
Since $z_1 z_2 > 0$ imposes the same condition as $\sign(z_1) = \sign(z_2)$ ($z_1, z_2 \in \R \setminus \{ 0 \}$), the standard estimator for $P(\diffxrv \diffyrv > 0)$ is
\begin{equation}
    \acc (\diffx, \diffy) \coloneqq \frac{\sum_{t \in \mathcal{T}} \ind{\diffxt \diffyt > 0}}{\card{T}}.\label{eq:acc}
\end{equation}
Here, the numerator counts the number of same-sign-changes, while the denominator is the number of considered pairs $(\diffyt, \diffxt)$.
Thus, $\acc$ is the proportion of concordant changes on all changes.
We refer to this estimator as the \textit{ATC ratio} of the prediction and set $\mathcal{T} = \{l, \dots, T\}$.
Visually, the measure computes the fraction of points in the upper right or lower left quadrant.
Similar evaluations are used in other scientific areas, for example, with contingency tables in dichotomous forecasting or with confusion matrices in classification analysis~\citetext{see, e.g., the introductions in \citealt{James2021}, Ch. 4, and \citealp{Jolliffe2012}, Ch. 3}.
Many other measures can be adapted from those fields to deepen the analysis.
Two simple measures that focus on a positive or negative predicted change are the positive and negative \ac{atc} ratios $\accp$ and $\accm$, respectively.
They are defined as
\begin{align}
    \accp (\diffx, \diffy) &\coloneqq \frac{\sum_{t \in \mathcal{T}} \ind{\diffxt \diffyt > 0} \ind{\diffxt > 0}}{\sum_{t \in \mathcal{T}} \ind{\diffxt > 0}},\ \text{and} \label{eq:accp}\\
    \accm (\diffx, \diffy) &\coloneqq \frac{\sum_{t \in \mathcal{T}} \ind{\diffxt \diffyt > 0} \ind{\diffxt < 0}}{\sum_{t \in \mathcal{T}} \ind{\diffxt < 0}}.\label{eq:accm}
\end{align}
They estimate the probability of a correct prediction of the direction of change, given that the predicted direction is positive or negative, that is, $P(\diffxrv \diffyrv > 0 | \diffxrv > 0)$ and $P(\diffxrv \diffyrv > 0 | \diffxrv < 0)$.

@@

\begin{figure}
    \centering
    \begin{subfigure}[t]{.24\textwidth}
\includegraphics{plots/illustrative_examples/4Q_sample_without_time}
\caption{Four-quadrant plot with simulated data.}\label{fig:aatc_basic_4q_sample}
\end{subfigure}\hspace{0.01\textwidth}
\begin{subfigure}[t]{.24\textwidth}
\includegraphics{plots/illustrative_examples/4Q_sample_with_time}
\caption{The same data is colored according to the time index $t$, the greener, the later.}\label{fig:aatc_basic_4q_sample_color}
\end{subfigure}\hspace{0.01\textwidth}
\begin{subfigure}[t]{.48\textwidth}
    \includegraphics{plots/illustrative_examples/trending_ratio_time_series}
    \caption{Rolling estimate of the \ac{atc} ratio over time with window length 91. }\label{fig:atc_ratio_time_series}
    \end{subfigure}%
    \caption[Visualizations of the four-quadrant plot for data with a time-varying ATC ratio.]{Visualizations for data with a time-varying ATC ratio. We defer information on the data generation process to the appendix (see \ref{subsec:app-ATC-data-generation}). The ATC ratio for the entire data set is $\mu = 0.7577$. The strong seasonality of the ATC ratio becomes visible in Figure~\ref{fig:atc_ratio_time_series}. The green curve $k_t$ shows the theoretical probability that $\diffxt$ has the same sign as $\diffyt$ for each time step. The rolling estimates are shifted to the right compared to $k_t$ as the windows look backward. Thus, the yearly course of the ATC ratio can be detected.
The ATC ratio has a pronounced sinus-shaped seasonality with a peak after a quarter of a year and a low point after three quarters.
}
\end{figure}

@@

The above measures can be extended to account for information on the point's location within the quadrant.
For example, points close to the zero point may have less explanatory power or may be less reliable than points far away from zero on one of the diagonals.
Suppose noise or non-systematic effects are present in the true values or predictions.
In that case, noise can drive a point's assignment to a quadrant instead of a systematic \ac{atc}.
This is more likely for points with at least one small coordinate.

Using an exclusion area around the zero point, as further defined below, is a straightforward and highly interpretable extension of the measures of Section~\ref{subsec:atc-measures} accounting for such effects~\citep[see, e.g.,][]{Saugel2015, Critchley2010}.
Points within that area are omitted in the calculation of the measures.
In particular, the \ac{mnf} are likely to have a noise component; thus, $\diffx$ should be subject to an exclusion area.
The measures of Equations~\eqref{eq:acc},~\eqref{eq:accp} and~\eqref{eq:accm} without points in the exclusion area $E$ are
\begin{align}
    \acceps (\diffx, \diffy, E) &\coloneqq \frac{\sum_{t \in \mathcal{T}} \ind{\diffx \diffy > 0} \ind{(\diffyt, \diffxt) \notin E}}{\sum_{t \in \mathcal{T}} \ind{(\diffyt, \diffxt) \notin E}},\label{eq:acceps}\\
    \accpeps (\diffx, \diffy, E) &\coloneqq \frac{\sum_{t \in \mathcal{T}} \ind{\diffxt \diffyt > 0} \ind{\diffxt > 0, (\diffyt, \diffxt) \notin E}}{\sum_{t \in \mathcal{T}} \ind{\diffxt > 0, (\diffyt, \diffxt) \notin E}}, \ \text{and} \label{eq:accpeps}\\
    \accmeps (\diffx, \diffy, E) &\coloneqq \frac{\sum_{t \in \mathcal{T}} \ind{\diffxt \diffyt > 0} \ind{\diffxt < 0, (\diffyt, \diffxt) \notin E}}{\sum_{t \in \mathcal{T}} \ind{\diffxt < 0, (\diffyt, \diffxt) \notin E}}. \label{eq:accmeps}
\end{align}
The measures are then estimators for the probability of predicting the correct direction, given that the point's location is not driven by noise or non-informative changes.

The estimators accept various shapes of the exclusion area (see Figure~\ref{fig:aatc_4q}).
A rectangular exclusion area, $E = \{(y, x) \in \R^2: (-\varepsilon_x \leq x \leq \varepsilon_x) \land (-\varepsilon_y \leq y \leq \varepsilon_y) \}$ for $\varepsilon_x, \varepsilon_y > 0$, leaves out points that are small in both components.
%Points are included if the true value or prediction is unlikely to be noise-driven.
%In the example graph in Figure~\ref{fig:aatc_basic_4q_excl_box}, only point 1 is excluded.
An exclusion area along one axis, for example, $E = \{(y, x) \in \R^2: (-\varepsilon_x \leq x \leq \varepsilon_x)\}$ for $\varepsilon_x > 0$, removes points in which one of the components could change sign by a small amount of noise.
%At least for \ac{mnf}, small amounts of noise are inevitable.
A cross-shaped exclusion area, $E = \{(y, x) \in \R^2: (-\varepsilon_x \leq x \leq \varepsilon_x) \lor (-\varepsilon_y \leq y \leq \varepsilon_y) \}$ for $\varepsilon_x, \varepsilon_y > 0$, along both axes accounts for the sign reversal in both components.
%A point is only included if no component is likely to be driven by noise.
%For the latter two methods, points 1 and 7 in Figure~\ref{fig:aatc_basic_4q_excl_axis} or 1, 6, and 7 are excluded in Figure~\ref{fig:aatc_basic_4q_excl_cross}, respectively.

@@

Confidence intervals can account for the estimation uncertainty of the measures above.
Bootstrap confidence intervals are based on resampling and not on parametric assumptions as classical confidence intervals are~\parencite[for introductions see][]{Hesterberg2011, Bittmann2021}.
Many new samples are drawn with replacement from the dataset, and the statistic of interest is computed for each sample, yielding an estimate for the distribution of the statistic of interest.
Based on the derived \enquote{new} samples of the statistic, the confidence intervals can be derived through different bootstrapping methods.
We use the \ac{bca} approach for bootstrapping in the following, as it holds the confidence level for small and large samples and has a moderate computation time (see the simulation study in Appendix~\ref{subsec:appendix-aatc-bootstrap}).

@@

The estimators described above give information on the probabilities $P(\diffxrv \diffyrv > 0 | \diffxrv \diffyrv \notin E)$, $P(\diffxrv \diffyrv > 0 | \diffxrv > 0, \diffxrv \diffyrv \notin E)$ and $P(\diffxrv \diffyrv > 0 | \diffxrv < 0, \diffxrv \diffyrv \notin E)$.
A still finer analysis might be gained by considering the conditional distribution $P(\diffxrv \diffyrv > 0 | \diffxrv = \xcond)$ to assess the \ac{atc} of a prediction for a specific change $\diffxrv = \xcond$ of the \ac{mnf}.
Thereby, $P(\diffxrv \diffyrv > 0 | \diffxrv = \xcond)$ denotes the probability of a correct direction given a predicted change of $\xcond$.
Thus, if a change of $\xcond$ is observed in practice, one can directly assess its credibility.
A multivariate \acf{kde} facilitates a continuous estimation of $P(\diffxrv \diffyrv > 0 | \diffxrv = \xcond)$ by estimating the components $f_{\diffxrv, \diffyrv}$ and $f_{\diffxrv}$ of
\begin{align*}
P(\diffxrv \diffyrv > 0 | \diffxrv = \xcond) = \begin{cases}
                                              \int_{-\infty}^0 \frac{f_{\diffxrv, \diffyrv}(\xcond, y)}{f_{\diffxrv}(\xcond)} \ \textrm{d} \: y & \text{if } \xcond < 0, \\
                                              \int_{0}^{\infty} \frac{f_{\diffxrv, \diffyrv}(\xcond, y)}{f_{\diffxrv}(\xcond)} \ \textrm{d} \: y & \text{if } \xcond > 0, \\
\end{cases}
\end{align*}
for $\xcond \neq 0$ through a \ac{kde}.
\textcite{Gramacki2018} provides a comprehensive introduction to multivariate \ac{kde}, and implementations are available in many programming languages, for example, in the \verb|statsmodels| in Python~\parencite{Seabold2010}.
The \ac{kde} yields estimates for $P(\diffxrv \diffyrv > 0 | \diffxrv = \xcond)$ for all values of $\xcond \in \R$.
Multivariate \ac{kde} takes a kernel and bandwidth selector as modeling parameters.
We advise using a Gaussian kernel and the cross-validation maximum likelihood as bandwidth selector (see Appendix~\ref{subsec:appendix-kde}).

Assessing $P(\diffxrv \diffyrv > 0 | \diffxrv = \xcond)$ graphically by drawing $P(\diffxrv \diffyrv > 0 | \diffxrv = \xcond)$ against $\xcond$ eases the simultaneous evaluation of various $\xcond$.
Furthermore, the graph facilitates the comparison of various methods in a single graph, and asymmetries of $P(\diffxrv \diffyrv > 0 | \diffx = \xcond)$ with respect to $\xcond$ in the \ac{atc} can be detected.
We refer to the plot as a \textit{conditional ATC plot}.

@@

In nowcasting and forecasting, probabilistic predictions have become more prevalent in recent years (see Sections~\ref{sec:application-covid} and~\ref{sec:application-eda}).
In this section, we develop \ac{atc} assessments for probabilistic measurements and nowcasts.
Probabilistic predictions issue a probability distribution for the quantity of interest based on their available information and, thus, include a point estimate and information on the prediction uncertainty and quantiles simultaneously.
Probabilistic predictions thus also contain a probability of a positive or negative change.
For \ac{atc} assessment, we compare the predicted probability of positive change, denoted by $p_t$, with the occurrence of positive changes.

Probabilistic predictions can be a \ac{cdf}, \ac{pdf}, or quantiles.
The \ac{cdf} is the most general and can be used to derive the others, given that they exist.
Let us first assume that the prediction is a \ac{cdf}, and that $y_{t-l}$ is known at time $t$ (see Table~\ref{tab:notation}).
Appendix~\ref{subsec:app-probabilistic-atc-evaluation} extends the analysis to quantile predictions or unknown true values.
